---
title: "Notebook 02: Graphics in R"
author: "Taylor Arnold and Lauren Tilton"
---

We need to reload the packages in each notebook. These are usually
included at the start of the notebook.

```{r, message = FALSE}
library(readr)               # read and write in the datasets
library(ggplot2)             # plotting library
library(dplyr)               # apply database-like commands
library(forcats)             # manipulate categorical variables
library(lubridate)           # work with dates and times

theme_set(theme_minimal())   # set a minimal plot as the default
options(
  dplyr.summarise.inform = FALSE,    # some additional options to reduce the
  readr.show_col_types = FALSE       # noise of some of the functions
)
```

## Data Visualization

### Bar Plots

For this notebook we will be using the photographs data. You will need to re-read
the data into R:

```{r, message = FALSE}
photos <- read_csv("data/photo_metadata_20200707.csv")    # read in the dataset
```

To start, we will build a bar plot showing how many photographs are taken in
each year. Try not too get to worried about the specific code syntax at this
point; instead, focus on what information is contained in the plot.

```{r, warning=FALSE}
photos %>%                  # built a plot from the photos data
  ggplot(aes(x = year)) +   # this says: start a plot with 'year' on the x-axis
    geom_bar()              # what kind of plot? a bar plot!
```

So far, so good. We can modify this code to instead count the number of photographs
taken by each photographer. This involves changing the just the variable mapped to
the x-axis.

```{r}
photos %>%                          # built a plot from the photos data
  ggplot(aes(x = photographer)) +   # this says: start a plot with 'photographer' on the x-axis
    geom_bar()                      # what kind of plot? a bar plot!
```

Unfortunately, this does not work so well. There are too many photographers and the plot
is impossible to read. To fix this, we can create a new variable that "lumps" together
photographers that did not take a large number of photographs. Here's the code:

```{r}
photos %>%                                   # built a plot from the photos data
  mutate(                                    # create a new variable 'p_new' that
    p_new = fct_lump(photographer, n = 16)   #   collapses all but the 16 photographers
  ) %>%                                      #   into a category called 'other'
  ggplot(aes(x = p_new)) +                   # this says: start a plot with 'p_new' on the x-axis
    geom_bar()                               # what kind of plot? a bar plot!
```

Starting to look better, but the labels are still too hard to read. We can fix this by
putting them on the y-axis instead of the x-axis:

```{r}
photos %>%                                   # built a plot from the photos data
  mutate(                                    # create a new variable 'p_new' that
    p_new = fct_lump(photographer, n = 16)   #   collapses all but the 16 photographers
  ) %>%                                      #   into a category called 'other'
  ggplot(aes(y = p_new)) +                   # this says: start a plot with 'p_new' on the y-axis
    geom_bar()                               # what kind of plot? a bar plot!
```

Great! Now we can actually read the plot and start the understand the data. It might
be better though to order to photographers by how many photographs the took. To do
this, we add another layer into the code that uses the function `fct_infreq`.

```{r}
photos %>%                                   # built a plot from the photos data
  mutate(                                    # create a new variable 'p_new' that
    p_new = fct_lump(photographer, n = 16)   #   collapses all but the 16 photographers
  ) %>%                                      #   into a category called 'other'
  mutate(p_new = fct_infreq(p_new)) %>%      # order the categories by frequency
  ggplot(aes(y = p_new)) +                   # this says: start a plot with 'p_new' on the y-axis
    geom_bar()                               # what kind of plot? a bar plot!
```

And as a final touch, we will get rid the "Other" and NA (missing) categories:

```{r}
photos %>%                                   # built a plot from the photos data
  mutate(                                    # create a new variable 'p_new' that
    p_new = fct_lump(photographer, n = 16)   #   collapses all but the 16 photographers
  ) %>%                                      #   into a category called 'other'
  mutate(p_new = fct_infreq(p_new)) %>%      # order the categories by frequency
  filter(!is.na(p_new)) %>%                  # removing missing photographer names
  filter(p_new != "Other") %>%               # remove the other category
  ggplot(aes(y = p_new)) +                   # this says: start a plot with 'p_new' on the y-axis
    geom_bar()                               # what kind of plot? a bar plot!
```

That looks good already for EDA. If we want to create a plot for publication, we
can add a few additional tweaks to make it look even better:

```{r}
photos %>%                                    # built a plot from the photos data
  mutate(                                     # create a new variable 'p_new' that
    p_new = fct_lump(photographer, n = 16)    #   collapses all but the 16 photographers
  ) %>%                                       #   into a category called 'other'
  mutate(p_new = fct_infreq(p_new)) %>%       # order the categories by frequency
  mutate(p_new = fct_rev(p_new)) %>%          # put photographer with most photos at the top
  filter(!is.na(p_new)) %>%                   # removing missing photographer names
  filter(p_new != "Other") %>%                # remove the other category
  ggplot(aes(y = p_new)) +                    # this says: start a plot with 'p_new' on the y-axis
    geom_bar(                                 # what kind of plot? a bar plot!
      fill = "white", color = "black"         #    with bars that are black outlines
    ) +                                       #    around a white center
    labs(                                     # labels for the plot
        x = "Number of Photographs",
        y = "Photographer",
        title = "Number of Digitzed Photographs Taken by FSA-OWI Photographers",
        caption = "Source: [photogrammar.org]"
     )
```

### Scatter Plots

Let's look at a few other plot types. For example, what if we want to look
at a count of two different varaibles? For this we might use a scatter plot,
also known as a "point" plot. Here's some code that tries to see the relationship
between year and month:

```{r, warning=FALSE}
photos %>%                     # built a plot from the photos data
  ggplot(aes(year, month)) +   # plot with 'year' on the x-axis and 'month' on y-axis
    geom_point()               # what kind of plot? a point plot!
```

As with our bar plot, the first pass does not look like much. All of the duplicate
points are stacked on one another and we cannot see any pattern. Let's change the
plot type to a count plot to fix this:

```{r, warning=FALSE}
photos %>%                          # built a plot from the photos data
  ggplot(aes(year, month)) +        # plot with 'year' on the x-axis and 'month' on y-axis
    geom_count() +                  # what kind of plot? a count plot!
    scale_size_area(max_size = 10)  # adjust the scale of the sizes
```

Much better! A remaining issue is that the axis scales are a bit off because
R thinks of year and month as continuous and not categorical. We can adjust
this with some preprocessing code:

```{r}
photos %>%                                  # built a plot from the photos data
  filter(!is.na(year), !is.na(month)) %>%   # removing missing values
  mutate(year = factor(year)) %>%           # make 'year' a categorical variable
  mutate(month = factor(month)) %>%         # make 'month' a categorical variable
  ggplot(aes(year, month)) +                # plot with 'year' on the x-axis and 'month' on y-axis
    geom_count() +                          # what kind of plot? a count plot!
    scale_size_area(max_size = 10)          # adjust the scale of the sizes
```

Or, perhaps even better, we can replace the numeric month with the month name:

```{r}
photos %>%                                 # built a plot from the photos data
  filter(!is.na(year), !is.na(month)) %>%  # removing missing values
  mutate(year = factor(year)) %>%          # make 'year' a categorical variable
  mutate(                                  #
    month = fct_inorder(month.abb[month])  # match 'month' with abbreviations
  ) %>%                                    #
  ggplot(aes(year, month)) +               # plot with 'year' on the x-axis and 'month' on y-axis
    geom_count() +                         # what kind of plot? a count plot!
    scale_size_area(max_size = 10)         # adjust the scale of the sizes
```

While not a focus of one of our topics today, R does have extensive support for plotting
dates as formal time series. Using the `make_date` function, we can show the number of
photographs taken in each month as a single plot:

```{r}
photos %>%                                         # built a plot from the photos data
  filter(!is.na(year), !is.na(month)) %>%          # removing missing values
  mutate(dt = make_date(year, month)) %>%          # make a date-time object
  group_by(dt) %>%                                 # group the data by photographer and date-time  
  summarize(n = n()) %>%                           # for each date-time, count number of photos
  ggplot(aes(dt, n)) +                             # plot with 'dt' on the x-axis and 'n' on y-axis
    geom_line()                                    # what kind of plot? a line plot!
```

### Adding Colors and Groups

A powerful feature of R graphics is that we can easily split a single plot
into groups by another feature (color, shape, size, ect.) accoring to
another variable.

```{r}
photos %>%                                         # built a plot from the photos data
  mutate(                                          # create a new variable 'p_new' that
    p_new = fct_lump(photographer, n = 10)         #   collapses all but the 5 photographers
  ) %>%                                            #   into a category called 'other'
  filter(!is.na(p_new)) %>%                        # removing missing photographer names
  filter(p_new != "Other") %>%                     # remove the other category
  filter(!is.na(year), !is.na(month)) %>%          # removing missing values
  mutate(year = make_date(year)) %>%               # make a date-time object
  group_by(p_new, year) %>%                        # group the data by photographer and date-time  
  summarize(n = n()) %>%                           # for each date-time, count number of photos
  ggplot(aes(year, n)) +                           # plot with 'year_month' on the x-axis and 'n' on y-axis
    geom_line(aes(color = p_new))                  # what kind of plot? a line plot! also, color by 'pnew'
```

### A Simple Spatial Plot

Finally, before getting to more formal spatial analysis later today, let's see how
we can use R graphics to plot the spatial component of the data. For example, a
simple map can be made as a scatter plot:

```{r, warning=FALSE}
photos %>%                                # built a plot from the photos data
  ggplot(aes(lon, lat)) +                 # plot with 'lon' on the x-axis and 'lat' on y-axis
    geom_point()                          # what kind of plot? a scatter plot!
```

The points in Alaska, Hawaii, Puerto Rico, and the Virgin Islands make the plot a bit hard to
see. Let's remove those for now.

```{r, warning=FALSE}
photos %>%                                                # built a plot from the photos data
  filter(between(lon, -150, 0), between(lat, 24,50)) %>%  # restrict the longitude and latitude
  ggplot(aes(lon, lat)) +                                 # plot with 'lon' on the x-axis and 'lat' on y-axis
    geom_point()                                          # what kind of plot? a scatter plot!
```

Seeing where photographs were taken is interesting, but perhaps we want to weight locations
based on the number of photos. We can do that by switching to a count plot. Also, black is a
bit boring, so we'll use a different color and make the points transparent:

```{r}
photos %>%                                                # built a plot from the photos data
  filter(between(lon, -150, 0), between(lat, 24,50)) %>%  # restrict the longitude and latitude
  ggplot(aes(lon, lat)) +                                 # plot with 'lon' on the x-axis and 'lat' on y-axis
    geom_count(                                           # what kind of plot? a count plot!
      color = "olivedrab",                                # color the points in a dark green color
      alpha = 0.2,                                        # make them transparent
      show.legend = FALSE                                 # and don't show a legend (to save room)
    ) +
    scale_size_area(max_size = 20) +                      # fix the scales to be proportional to area
    theme_void()                                          # remove all the cluter to focus on the data
```

We can learn a lot about the data from small modifications to this plot. For example,
what if we only take photos from a single photographer? We could discover where they
took photos relative to the other photographers. Here's a plot showing the route of
John Vachon:

```{r}
photos %>%                                                # built a plot from the photos data
  filter(between(lon, -150, 0), between(lat, 24,50)) %>%  # restrict the longitude and latitude
  filter(photographer == "John Vachon") %>%               # only take a single photographer
  ggplot(aes(lon, lat)) +                                 # plot with 'lon' on the x-axis and 'lat' on y-axis
    geom_count(                                           # what kind of plot? a count plot!
      color = "olivedrab",                                # color the points in a dark green color
      alpha = 0.2,                                        # make them transparent
      show.legend = FALSE                                 # and don't show a legend (to save room)
    ) +
    scale_size_area(max_size = 20) +                      # fix the scales to be proportional to area
    theme_void()                                          # remove all the cluter to focus on the data
```

Because Vachon did not take photos everywhere, it's a bit harder to see exactly where all
the points would line up on a map. We will see a way of fixing this in a following set of
notes.

### On Your Own

Pick a different photographer (use the earlier plots to determine some good candidates)
and create a similar plot to the one we created with John Vachon. Feel free to choose a
different color for the points (see http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf
for available color names; hex codes work too).

```{r}

```

For an added challenge, try to filter your plot by a specific range of years.
